---
title: "R Notebook"
output:
  word_document: default
  html_notebook: default
---
#BIS 581
##Paul Dunn
###Missing Value imputation

The code below uses random forests (covered later, don't worry about them for now) to fill in some missing values (na) in a data set.
The iris dataset is built-in and has data about flowers in it. Feel free to look at the dataset and/or open the help file for it.

```{r}
data(iris)
#let's get our own copy
iris.na <- iris
set.seed(111)
## artificially drop some data values.
for (i in 1:4) iris.na[sample(150, 20), i] <- NA

#how many missing values do we have?
summary(iris.na)
View(iris)
View(iris.na)

```

This shows us how many missing values we have, spread across the data set.
Just for comparison's sake, let's look at an original value, the row where that value is missing and then the value our function (below) fills in
```{r}
iris[6,4]
iris.na[6,4]
```

won't be perfect, but let's fill in the values with impute, randomforest
```{r}
library(randomForest)
set.seed(222)
iris.imputed <- rfImpute(Species ~ ., iris.na)
View(iris.imputed)
```
OK, now let's compare values again
```{r}
iris[3,2]
iris.na[3,2]
iris.imputed[3,2]
print(1-iris.imputed[3,2]/iris[3,2])
```

Not perfect by any means, but if we need all of the rows, it works. Take a look at the last line printed above. this is just one minus the ratio of the imputed value to the 'real' value. 