---
title: "R Notebook"
output:
  word_document: default
  html_notebook: default
---
BIS581
This assignment will look at some stock market data. Answer each question, provide your r code also. 
You will need to edit the line below to point it at where you saved the CSV file

```{r}
my_stocks <- read.csv("stocks.csv",header=TRUE, stringsAsFactors = FALSE)
```

Your job first, is to 'know your data'. Run whatever code you feel necessary to get to know your data. Add more code blocks as needed, leave this part in your document when you submit it.

```{r}
str(my_stocks)

```

For the questions below, show both the code used to answer along with text that shows the answer: 

How many different stocks are there (by stock symbol) (5)?
```{r}
# Get unique stock symbols
unique_stocks <- unique(my_stocks$stock_symbol)

# Count how many unique stock symbols there are
num_unique_stocks <- length(unique_stocks)

#  result
num_unique_stocks
```

What are those stocks (5)?
```{r}
# Get unique stock symbols
unique_stocks <- unique(my_stocks$stock_symbol)

# results
unique_stocks
```

Which stock had the highest single gain in a day, which day was it (10)?
```{r}
my_stocks$daily_gain <- my_stocks$stock_price_close - my_stocks$stock_price_open

# Find the stock with the maximum daily gain
max_gain_row <- my_stocks[which.max(my_stocks$daily_gain), ]

# result
max_gain_row$stock_symbol
max_gain_row$date
max_gain_row$daily_gain
```

Which stock had the highest single loss in a day, which day was it (10)?
```{r}
# Calculate daily gain
my_stocks$daily_gain <- my_stocks$stock_price_close - my_stocks$stock_price_open

# Find the stock with the largest single loss (minimum daily gain)
min_loss_row <-my_stocks[which.min(my_stocks$daily_gain), ]

#  result
min_loss_row$stock_symbol
min_loss_row$date
min_loss_row$daily_gain
```

Create a plot of closing price for XOM over time (5)
```{r}
# Convert date column to Date type
my_stocks$date <- as.Date(my_stocks$date)

# Filter the data for XOM stock symbol
xom_data <- subset(my_stocks, stock_symbol == "XOM")

# Plot closing price over time
plot(xom_data$date, xom_data$stock_price_close, type = "l", col = "blue", 
     main = "XOM Closing Price Over Time", xlab = "Date", ylab = "Closing Price")
```

Create a plot of closing price for XKK over time (5)
```{r}
# Convert date column to Date type
my_stocks$date <- as.Date(my_stocks$date)

# Filter the data for XKK stock symbol
xkk_data <- subset(my_stocks, stock_symbol == "XKK")

# Plot closing price over time
plot(xkk_data$date, xkk_data$stock_price_close, type = "l", col = "red", 
     main = "XKK Closing Price Over Time", xlab = "Date", ylab = "Closing Price")

```

Create a plot of the change in price per day for both XOM and XKK (10)
```{r}
# Calculate daily price change
my_stocks$daily_change <- my_stocks$stock_price_close - my_stocks$stock_price_open

# Filter data for XOM and XKK
xom_data <- subset(my_stocks, stock_symbol == "XOM")
xkk_data <- subset(my_stocks, stock_symbol == "XKK")

# Plot for XOM
plot(xom_data$date, xom_data$daily_change, type = "l", col = "blue", ylim = range(my_stocks$daily_change),
     main = "Daily Price Change for XOM and XKK", xlab = "Date", ylab = "Daily Change")

```

